---
title: "Miscellaneous Utilities"
author: "Nicholas Mikolajewicz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_chunk$set(
  collapse = F,
  comment = "#>"
)

# clear global enviroment 
rm(list = ls())

# load XcalRep
library(XcalRep)

```

# Miscellaneous Functions

There are several additional functions that are worthwhile exploring in the `XcalRep` package.

### Data import and prep

Prior to analysis, import data and preprocess as described in the **Getting Started** Vignette. 

```{r data prep, warning = F, message = F}

# import EFP and QC1 imaging phantom data
df.qc1 <-qc1
df.efp <- efp

# omit water-mimetic resin sections
df.qc1 <- dplyr::filter(df.qc1, section  != 1) # remove water mimic section 1
df.efp <- dplyr::filter(df.efp, section  != 4) # remove water mimic section 4
df.efp$section <- as.numeric(as.character(df.efp$section))
df.qc1$section <- as.numeric(as.character(df.qc1$section))

# combine datasets (ensure replicate sets aren't mixed)
df.qc1$replicateSet <- paste("q", df.qc1$replicateSet, sep = "")
df.efp$replicateSet <- paste("e", df.efp$replicateSet, sep = "")
all.data <- bind_rows(df.qc1, df.efp)


# create Calibration Object
co <- createCalibrationObject(all.data)

# get list of feature subsets to analyze
analyze.these <- analyzeWhich(co, 
                              include.parameters = c("Tt.vBMD"))

# preprocess/filter data
co <- preprocessData(object = co, 
                      analyze.which = analyze.these, 
                      new.assay.name = "preprocessed.data",
                      which.assay = "input")

```


**1) Mean-Variance Relationship:** `meanVarPlot`

The `mean.var.plot` function enables users to explore the relationship between measurement means and variance, and may inform which precision error is more appropriate to report. The precision error measure that remains relatively constant across all values is preferred, as it does not misrepresent reproducibility for extreme values. 

This relationship can be examined as either a **scatter plot**

```{r meanVarPlot scatter}
meanVarPlot(co, 
              which.data = "uncalibrated", 
              which.plot = "scatter", 
              which.parameter = "Tt.vBMD")
```

or **boxplot**

```{r meanVarPlot box, fig.width= 10, fig.height= 3.5}

# requires troubleshooting
# error message: Error in eval(predvars, data, env) : object 'section' not found

# meanVarPlot(co, 
#               which.data = "uncalibrated", 
#               which.plot = "box", 
#               which.parameter = "Tt.vBMD")
```


**2) Utility Functions:** 

`cloneAssay`
If users i) create an analysis "checkpoint" to which they can return at a later time or ii) diverge in analysis pipelines for the current dataset, `cloneAssay` will duplicate the specified (or current) `Assay` within the current `Calibration Object`.  

```{r clone assay}
# duplicate current assay
co <- cloneAssay(co)
getAssay(co, which.assays = "all")
```



`deleteAssay`
A specified assay can be deleted using the `deleteAssay` function. 

```{r delete assay}
# duplicate current assay
co <- deleteAssay(co, which.assay = "preprocessed.data-copy")
```



