---
title: "Cross-Calibration"
author: "Nicholas Mikolajewicz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_chunk$set(
  collapse = F,
  comment = "#>"
)

# clear global enviroment 
rm(list = ls())

# load XcalRep
library(XcalRep)

```

# Cross-Calibration 

Imaging phantoms are used to cross-calibrate instruments/scanners thereby enabling direct comparion of measurements obtained by different instruments. In general, a reference instrument is selected, and pair-wise linear regression of imaging phantom measurements is used to calibrate all other instruments to the reference. For HR-pQCT, imaging phantoms (e.g., QC1, EFP) are designed to mimic varying degrees of bone mineral densities, along with certain other properties, thereby allowing calibration using multiple points of reference. 

### Data import and prep

Prior to analysis, import data and preprocess as described in the **Getting Started** Vignette. 

```{r data prep, warning = F, message = F}

# import EFP and QC1 imaging phantom data
df.qc1 <-qc1
df.efp <- efp

# omit water-mimetic resin sections
df.qc1 <- dplyr::filter(df.qc1, section  != 1) # remove water mimic section 1
df.efp <- dplyr::filter(df.efp, section  != 4) # remove water mimic section 4
df.efp$section <- as.numeric(as.character(df.efp$section))
df.qc1$section <- as.numeric(as.character(df.qc1$section))

# combine datasets (ensure replicate sets aren't mixed)
df.qc1$replicateSet <- paste("q", df.qc1$replicateSet, sep = "")
df.efp$replicateSet <- paste("e", df.efp$replicateSet, sep = "")
all.data <- bind_rows(df.qc1, df.efp)


# create Calibration Object
co <- createCalibrationObject(all.data)

# get list of feature subsets to analyze
analyze.these <- analyzeWhich(co, 
                              include.parameters = c("Tt.vBMD"))

# preprocess/filter data
co <- preprocessData(object = co, 
                      analyze.which = analyze.these, 
                      new.assay.name = "preprocessed.data",
                      which.assay = "input")

```

### Identify reference

Before we fit pair-wise calibration curves, we must first specify our point of reference with respect to the instrument (denoted `reference.site`) and time (if scans were obtained at multiple timepoints). 

We propose designating the reference site as the site that reports the most precise and accurate measurements. For a detailed description of how most precise and accurate site is derived, refer to our companion paper (**citation placeholder**). For the reference time, we recommend using baseline (t=0) measurements. In general, it is good practice to provide justification for the choice of reference. 

Using our running example, phantom measurements obtained by the "Oregon" scanner were the most precise and accurate. 

```{r reference site, warning = F, message = F}
reference.site <- identifyReference(co)
reference.site
```

Using the `consistencyPlot` function, we can explore how each site ranked across all parameters. Ranking is based on a minimized MSE-criteria, computed as the mean-squared error (MSE) between site-specific measures and the median measurement values pooled across all `timePoint` and `site` features. The `reference.site` is then selected as the site which has the highest average ranking across all parameters. The `tile` plot enables us to examine how each site ranked in a parameter-specific manner. 

```{r consistency tile, fig.width = 9, fig.height = 4, warning = F, message=F}

# generate site ranking plots
qc1.c.plt <- consistencyPlot(co, which.plot = "tile", var2plot = "mse", which.phantom = "QC1")
qc1.c.plt <- qc1.c.plt +labs(subtitle = "QC1")

efp.c.plt <- consistencyPlot(co, which.plot = "tile", var2plot = "mse", which.phantom = "EFP")
efp.c.plt <- efp.c.plt +labs(subtitle = "EFP")

# combine plots
cowplot::plot_grid(qc1.c.plt, efp.c.plt)

```

To retrieve the scores, we can call the same funciton `identifyReference` and indicate that we want scores with `return.scores = T`

```{r}

# specify parameter of interest
par4study <- "Tt.vBMD"

# get reference scores
reference.scores <- identifyReference(co,
                                    which.phantom = c("EFP", "QC1"),
                                    which.parameter = par4study,
                                    pool.phantoms = T,
                                    return.scores = T)

# print data.table
(reference.scores[["QC1"]]) %>%
  datatable(extensions = 'Buttons', options = list(
     dom = 'Bfrtip',
     buttons = c('copy', 'csv', 'excel', 'pdf', 'print')) )

```

### Fit calibration curves

To fit separate calibration curves to the two different imaging phantoms (QC1 and EFP), first clone the preprocessed assays so that they can be calibrated in parallel. 

```{r clone assays}

# check which assays are available
getAssay(co, which.assays = "all")

# clone 'preprocessed.data' assay, once for each phantom that will be used for calibration
co <- cloneAssay(co, which.assay = "preprocessed.data", cloned.assay.name = "EFP")
co <- cloneAssay(co, which.assay = "preprocessed.data", cloned.assay.name = "QC1")

# check that new assays are available in calibration object
getAssay(co, which.assays = "all")
```

Compute calibration equations for the designated reference site using `fitCalibration`. Note that we specify 'Oregon' as the reference site. 


```{r fit calibration, warning = F, message=F}

# fit calibration curves using EFP imaging phantom
co <- fitCalibration(co, 
                     which.assay = "EFP",
                     reference.site = "Oregon",
                     which.phantom = "EFP",
                     sig.intercept.only = F, which.center = "mean")

getCalibrationInfo(co, which.assay = "EFP")

# fit calibration curves using QC1 imaging phantom
co <- fitCalibration(co, 
                     which.assay = "QC1",
                     reference.site = "Oregon",
                     which.phantom = "QC1",
                     sig.intercept.only = F,  which.center = "mean")

getCalibrationInfo(co, which.assay = "QC1")

```
*Tip:* Cross-calibration requires atleast 3 calibration points in each phantom. If less than 3 imaging phantom sections are detected in the data, an error will occur notifying the user. 

### Calibration equations

Once calibration equation have been computed, we can retrieve the results from the `Calibration Object` using `getResults`. Results can be directly exported to a spreadsheet from the interactive data table. Continuing with our example, we can examine the calibration curves fit for each imaging phantom.  
```{r get calibration results, warning = F, message=F}

# retrieve calibration results (as datatables)
calibration.results.qc1 <-getResults(object = co,
                                which.results = "calibration", # get calibrated data
                                which.assay = "QC1",           # QC1 calibrations
                                format = 'df')                 # retrieve as data.frame

calibration.results.efp <-getResults(object = co,
                                which.results = "calibration", # get calibrated data
                                which.assay = "EFP",           # EFP calibrations
                                format = 'df')                 # retrieve as data.frame


# see what tables were generated
names(calibration.results.qc1)
names(calibration.results.efp)

# show calibration equations for each imaging phantom
showTable(calibration.results.qc1[["calibration.equations"]], as.dt = T)
showTable(calibration.results.efp[["calibration.equations"]], as.dt = T)

```

### Visualization

To visualize the calibration curves, use the `calibrationPlot` function. 
```{r calibration curves, fig.width = 7, fig.height = 5}

calibrationPlot(co,
                which.assay = "QC1",
                 which.parameter = "Tt.vBMD",
                 return.plt.handle = F,
                 color.option = "A", color.begin = 0, color.end = 0.5)

calibrationPlot(co,
                which.assay = "EFP",
                 which.parameter = "Tt.vBMD",
                 return.plt.handle = F,
                 color.option = "A", color.begin = 0, color.end = 0.5)
```

### Data calibration

To calibrate phantom data using the computed calibration equations, use the `calibrateData` function. This will generate a new dataset called `calibrated.data` within the specified `Assay`. 

```{r calibrate data}
# calibrate data
co <- calibrateData(co, which.assay = "QC1", verbose = F)
co <- calibrateData(co, which.assay = "EFP", verbose = F)
```

The current `Assay` now contains two datasets; `uncalibrated` and `calibrated`.
```{r get data calibration}
# show available datasets
getDatasets(co, which.assay = "QC1")
getDatasets(co, which.assay = "EFP")
```

### Diagnostics

A **diagnostic plot** comparing pre- and post-calibrated values is useful in determining whether all sites were successfully calibrated. XcalRep enables users to generate three kinds of diagnostic plots using the `diagnosticPlot` function. 

**1) Diagnostic Line Plot:** 
Pre- and post-calibration values are plotted on the x and y axes, respectively, and 2 sets of reference curves are overlaid; Horizontal dashed references denote section-specific reference values which were used for calibration. The diagonal dashed reference is the line of equality (x = y) and the reference site lies along this curve.  

```{r diagnostic line, fig.width= 7, fig.height = 4}
diagnosticPlot(co, which.assay = "QC1", which.parameter = "Tt.vBMD", which.plot = "line")
```

**2) Diagnostic Bar Plot:** 
Pre- and post-calibration values are plotted on a horizontal bar plot, and the median value is shown as a dashed reference. 

```{r diagnostic bar, fig.width= 7, fig.height = 6, message = F}
diagnosticPlot(co, which.assay = "EFP", which.parameter = "Tt.vBMD", which.plot = "bar", fix.axis = T)
```

**3) Diagnostic Residual Distributions:** 
The distribution of pre- and post-calibration residuals is visualized using a histoogram and overlaid density curve. Residuals are here defined as the difference between the replicate mean and pooled median. 

```{r residuals, fig.width= 7, fig.height = 5, message = F}
diagnosticPlot(co, which.assay = "QC1", which.parameter = "Tt.vBMD", which.plot = "residual", fix.axis = T)
```


