---
title: "Multi-Variant Precision Analysis"
author: "Nicholas Mikolajewicz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_chunk$set(
  collapse = F,
  comment = "#>"
)

# clear global enviroment 
rm(list = ls())

# load XcalRep
library(XcalRep)

```

# Multi-Variant Precision Analyses

**Multi-variant precision (MVP)** analysis aims to calculate the inherent variability of an instrument and additional error arising from time- or scanner-dependent drifts in scanner calibration. MVP analysis is commonly referred to as **multisite-longitudinal precision** or reproducibility analysis and is necessary to perform when conducting a multisite or longitudinal clincial trial involving patient scans over multiple scanners and timepoints. 

### Data import and prep

Prior to analysis, import data and preprocess as described in the **Getting Started** Vignette. 

```{r data prep, warning = F, message = F}

# import EFP and QC1 imaging phantom data
df.qc1 <-qc1
df.efp <- efp

# omit water-mimetic resin sections
df.qc1 <- dplyr::filter(df.qc1, section  != 1) # remove water mimic section 1
df.efp <- dplyr::filter(df.efp, section  != 4) # remove water mimic section 4
df.efp$section <- as.numeric(as.character(df.efp$section))
df.qc1$section <- as.numeric(as.character(df.qc1$section))

# combine datasets (ensure replicate sets aren't mixed)
df.qc1$replicateSet <- paste("q", df.qc1$replicateSet, sep = "")
df.efp$replicateSet <- paste("e", df.efp$replicateSet, sep = "")
all.data <- bind_rows(df.qc1, df.efp)


# create Calibration Object
co <- createCalibrationObject(all.data)

# get list of feature subsets to analyze
analyze.these <- analyzeWhich(co, 
                              include.parameters = c("Tt.vBMD", "Tb.vBMD"))

# preprocess/filter data
co <- preprocessData(object = co, 
                      analyze.which = analyze.these, 
                      new.assay.name = "preprocessed.data",
                      which.assay = "input")

```

### 6.1 MVP Analysis

Run `mvpAnalysis` to perform MVP analysis and store results in `Calibration Object`. By specifying which.data = "all", MVP analysis will be performed on all data in the current `Assay`, and stratified by calibration (uncalibrated vs. calibrated) status.

```{r mvp analysis}
co <- mvpAnalysis(co, 
                   which.data = "all",  
                   verbose = F)


# retrieve stored results (as datatables)
mvp.results <-getResults(object = co, 
                                which.results = "mvp",
                                format = 'dt') 


# show rms statistics for calibrated data
showTable(mvp.results[["rms.statistics"]])
```
Alternatively, we can focus on just multisite longitudinal precision error

```{r}
# retrieve stored results (as data.frame)
mvp.results <-getResults(object = co,
                                which.results = "mvp",
                                format = 'df')
# get rms statistics
mvp.rms <- mvp.results[["rms.statistics"]]
# subset rms statistics
mvp.rms.subset <- mvp.rms[grepl(c("long-t12.multi"), mvp.rms$precision.type),  ]
# show table
showTable(mvp.rms.subset, as.dt = T)
```


### 4.3 Visualize Results

To visualize `mvpAnalysis` results, use the `mvpPlot` function. Function arguments are similar to those in the `svpPlot` function, with the additional option to subselect different precision error types, including:

- **short**: Short-term precision errors only

- **long**: Longitudinal precision errors (i.e., long-term) only

- **single**: Single-site precision errors only

- **multi**: Multi-site precision errors only

- **all**: all of the above (default)


```{r mvp plot}
mvpPlot(co, 
         outliers = T, 
         var2plot = "cv",
         which.data = "all", 
         which.precision = "all",
         color.begin = 0.1, 
         color.end = 0.5,
         jitter.width = 0.1,
         show.rms.statistic = T)
```

